---
title: "DATA 556 Homework 6 Lab"
author: "Will Wright"
date: "November 6, 2018"
output: pdf_document
---

```{r setup, message=FALSE, warning=FALSE}
library(kableExtra)
library(dplyr)
```

## Problem 6

Let X and Y be i.i.d. Unif(0,1).
a) Use simulations in R to numerically estimate the covariance of X + Y and X - Y.

```{r 6a}
set.seed(0)
n <- 10000
X <- runif(n)
Y <- runif(n)
X_plus_Y <- X + Y
X_minus_Y <- X - Y
cov(X_plus_Y,X_minus_Y)
```

## Problem 8 Setup

```{r 8}
y=matrix(c(0.018, 0.035, 0.031, 0.008, 0.018,
           0.002, 0.112, 0.064, 0.032, 0.069,
           0.001, 0.066, 0.094, 0.032, 0.084,
           0.001, 0.018, 0.019, 0.010, 0.051,
           0.001, 0.029, 0.032, 0.043, 0.130) , nrow =5 , byrow=TRUE)
colnames(y) =c("farm", "operatives", "craftsmen", "sales", "professional")
rownames(y) =colnames(y)

```


## Problem 8a
Using this joint distribution, calculate the marginal probability distribution of a father's occupation

```{r 8a}
kable(margin.table(y,1)) %>% kable_styling()
```

## Problem 8b
Using this joint distribution, calculate the marginal probability distribution of a son's occupation

```{r 8b}
kable(margin.table(y,2)) %>% kable_styling()
```

## Problem 8c
Using this joint distribution, calculate the conditional distribution of a son's occupation, given that the father is a farmer

```{r 8c}
kable(y[1,]/sum(y[1,])) %>% kable_styling()
```


## Problem 8d
Using this joint distribution, calculate the conditional distribution of a father's occupation, given that the son is a farmer


```{r 8d}
kable(y[,1]/sum(y[,1])) %>% kable_styling()
```

