---
title: "DATA 556: Homework 4"
author: "Will Wright"
date: "October 25, 2018"
output: pdf_document
---

```{r setup}
# load packages
library(ggplot2)
set.seed(0) # set seed for reproducibility
```

## General note
The encoding of certain math symbols like theta are not working as expected and display as '??'.  This, however, will have no impact on the answers--just the quoated questions.

## Problem 2a

A circle with a random radium R ~ Unif(0,1) is generated. Let A be its area.
a.) Use simulations in R (the statistical programming language) to numerically estimate the mean and variance of A.


```{r 2a}
circleArea <- pi*runif(10000)^2 # simulate 10,000 circle areas
mean(circleArea)
var(circleArea)
```

## Problem 3a

A stick of length 1 is broken at a uniformly random point, yielding two pieces.  Let X and Y be the lengths of the shorter and longer pieces, respectively, and let R = X/Y be the ratio of the lengths X and Y.
a.) Use simulations in R (the statistical programming language) to gain some understanding about the distribution of the random variable R.  Numerically estimate the expected value of R and 1/R.


```{r 3a, fig.height=9, fig.width=8}
stickBreakPoints <- runif(10000) # simulate 10,000 stick breaks
# create empty dataframe to hold the X and Y lengths
xyLengths <- setNames(data.frame(matrix(nrow = length(stickBreakPoints), ncol = 2)), c("X","Y"))

# loop to assign X and Y for each breakpoint, based on which is longer
for(i in 1:length(stickBreakPoints)){
  ifelse(stickBreakPoints[i]>=0.5,
         xyLengths$Y[i] <- stickBreakPoints[i],
         xyLengths$X[i] <- stickBreakPoints[i])
}

# update the NAs with 1-(the other value)
xyLengths$X[which(is.na(xyLengths$X))] <- 1-xyLengths$Y[which(is.na(xyLengths$X))]
xyLengths$Y[which(is.na(xyLengths$Y))] <- 1-xyLengths$X[which(is.na(xyLengths$Y))]

# Calculate R and 1/R
R <- xyLengths$X/xyLengths$Y
Rreciprocal <- 1/R

par(mfrow = c(2,2))
hist(R)
# plot without the outliers to better see the distribution
hist(Rreciprocal[-which(Rreciprocal %in% boxplot.stats(Rreciprocal)$out)], 
     main = "Histogram of 1/R",
     xlab = "1/R")
boxplot(R, main = "Boxplot of R")
boxplot(Rreciprocal, main = "Boxplot of 1/R")

# calculate E(R) and E(1/R) via their means and show summary stats
summary(R)
summary(Rreciprocal)
```

## Problem 4c

Use simulations in R to numerically estimate E(X).

```{r 4c, message = FALSE, warning = FALSE}
# simulations for n = 1 to n = 1000
n <- 1:1000
joinedMaxRandoms <- c()

for(i in 1:length(n)){
  randoms <- runif(i)
  maxRandom <- max(randoms)
  joinedMaxRandoms <- c(joinedMaxRandoms, maxRandom)
}

mean(joinedMaxRandoms) # this is a bit misleading since n varies with this approach
# show convergence to n/(n+1) via plot
g <- ggplot(data.frame(joinedMaxRandoms), aes(x = n, y = joinedMaxRandoms))
g + geom_point() +
  geom_smooth() +
  ylim(0.9,1) +
  theme_bw()

# I can show for a particular n, but I'm hoping this is more interesting?
```


## Problem 5b

Use simulations in R (the statistical programming language) to numerically estimate P(X < Y) for
X ~ N(0, 1), Y ~ N(1, 5) with X and Y independent.

```{r 5b}
n <- 10000 #sample size

# draw samples for both X and Y
X <- rnorm(n)
Y <- rnorm(n, mean = 1, sd = sqrt(5))

# calculate how many times X < Y
count <- sum(X<Y)
# calculate the probability by dividing the cases where X < Y by n
prob <- count/n
prob

```


## Problem 6b

Using simulations in R, calculate the Monte Carlo estimates of the mean and standard deviation of
the distribution of x ~ y. Compare these estimates with the exact values of the mean and standard deviation.

```{r 6b}
# create empty vars to store the means from the simulations
maleMeans <- c()
femaleMeans <- c()
deltaMeans <- c()

# loop to do simulations
for(i in 1:500){
  x <- rnorm(1:100, 69.1, 2.9)
  y <- rnorm(1:100, 63.7, 2.7)
  maleMeans[i] = mean(x)
  femaleMeans[i] = mean(y)
  deltaMeans[i] = maleMeans[i]-femaleMeans[i]
}

mean(deltaMeans)
sd(deltaMeans)
```


## Problem 6c

What is the probability that a randomly sampled man is taller than a randomly sampled woman?
Please do not answer this question by reporting a Monte Carlo estimate of this probability.

```{r 6c}
# calculate the exact probability
dist <- (69.1-63.7)/sqrt(2.9^2 + 2.7^2)
pnorm(dist, 0, 1)
```

## Problem 7b

For y = 0, make a plot for P(?? | y) for each ?? ??? ?? = {0.0, 0.1, . . . , 0.9, 1.0}. In other words, make
a plot with the horizontal axis representing the 11 values of ?? and the vertical axis representing the
corresponding values of P(?? | y).

```{r 7b}
# vector to store theta values
thetaVals <- seq(0.0, 1.0, by = 0.1)

# function to evaluate probability of theta given Y
probabilizer <- function(Y) {
  numerators <- c()
  
  # use equation from 7a
  for(i in 1:length(thetaVals)) {
    numerators <- c(numerators, ((thetaVals[i]^Y)*(1 - thetaVals[i])^(5-Y)))
  }
  
  denominator <- sum(numerators)
  
  probs <- c()
  for(i in 1:length(thetaVals)) {
    probs <- c(probs, (numerators[i] / denominator))
  }
  return(probs)
}

# calculate the probs for y = 0 
yCoords <- probabilizer(0)

# convert to data.frame for ggplot
plotData <- data.frame(x = thetaVals, y = yCoords)

g <- ggplot(plotData, aes(x = x, y = y))
g + geom_line() +
  theme_bw() +
  labs(x = "Theta Values", y = "Probability")
```


## Problem 7c

Repeat (b) for each y ??? {1, 2, 3, 4, 5}, so in the end you have six plots (including the one in (b)).
Describe what you see in your plots and discuss whether or not they make sense.

```{r 7c, fig.width = 8, fig.height=8}
# build a dataframe that can be used with facet_wrap() by making a tall dataset
# this is terrible code, but I'm short on time so beauty is out the window!!
plotData <- data.frame(x = rep(thetaVals, 5), 
                       y = c(probabilizer(1), 
                             probabilizer(2),
                             probabilizer(3),
                             probabilizer(4),
                             probabilizer(5)),
                       givenY = as.character(c(rep(1,11),rep(2,11),rep(3,11),rep(4,11),rep(5,11))))

g <- ggplot(plotData, aes(x = x, y = y))
g + geom_line() +
  theme_bw() +
  labs(x = "Theta Values", y = "Probability") +
  facet_grid(givenY ~ .)



```

## Problem 8b
 Implement your sampling algorithm in R, and use your code to produce a Monte Carlo estimate of
P(X ??? (2, 3)) where X is a random variable that has a Logistic distribution. 

```{r 8b}
n <- 10000 #simulations
x <- runif(n, 0, 1) # build vector of random uniforms
y <- -log((1/x)-1) # equation from 8a

successes <- sum(2<y & y<3)
prob <- successes/n
prob

```










