---
title: "DATA 556: Homework 4"
author: "Will Wright"
date: "October 25, 2018"
output: pdf_document
---

```{r setup}
# load packages
library(ggplot2)
set.seed(0) # set seed for reproducibility
```

## Problem 2a

A circle with a random radium R ~ Unif(0,1) is generated. Let A be its area.
a.) Use simulations in R (the statistical programming language) to numerically estimate the mean and variance of A.


```{r 2a}
circleArea <- pi*runif(10000)^2 # simulate 10,000 circle areas
mean(circleArea)
var(circleArea)

```

## Problem 3a

A stick of length 1 is broken at a uniformly random point, yielding two pieces.  Let X and Y be the lengths of the shorter and longer pieces, respectively, and let R = X/Y be the ratio of the lengths X and Y.
a.) Use simulations in R (the statistical programming language) to gain some understanding about the distribution of the random variable R.  Numerically estimate the expected value of R and 1/R.

```{r 3a}
stickBreakPoints <- runif(10000) # simulate 10,000 stick breaks
# create empty dataframe to hold the X and Y lengths
xyLengths <- setNames(data.frame(matrix(nrow = length(stickBreakPoints), ncol = 2)), c("X","Y"))

# loop to assign X and Y for each breakpoint, based on which is longer
for(i in 1:length(stickBreakPoints)){
  ifelse(stickBreakPoints[i]>=0.5,
         xyLengths$Y[i] <- stickBreakPoints[i],
         xyLengths$X[i] <- stickBreakPoints[i])
}

# update the NAs with 1-(the other value)
xyLengths$X[which(is.na(xyLengths$X))] <- 1-xyLengths$Y[which(is.na(xyLengths$X))]
xyLengths$Y[which(is.na(xyLengths$Y))] <- 1-xyLengths$X[which(is.na(xyLengths$Y))]

# Calculate R and 1/R
R <- xyLengths$X/xyLengths$Y
Rreciprocal <- 1/R

par(mfrow = c(2,2))
hist(R)
# plot without the outliers to better see the distribution
hist(Rreciprocal[-which(Rreciprocal %in% boxplot.stats(Rreciprocal)$out)])
boxplot(R)
boxplot(Rreciprocal)

# calculate E(R) and E(1/R) via their means and show summary stats
summary(R)
summary(Rreciprocal)
```
